---
title: "Module 2 assignment -  Statistical Analysis Presentation"
author: "Guilherme Amorim"
date: "2024-10-17"
output: pdf_document
---

# Initial setup

```{r include=FALSE}

knitr::opts_chunk$set(warning=FALSE, message = FALSE)


```

## Load libraries
```{r, message=F}

library(haven) # load .sav files

library(tidyverse) # data manipulation and visualisation

library(magrittr) # pipe operator for data manipulation

library(patchwork) # merging plots

library(DescTools) # pre-built function to calculate mode

library(gmodels) # contigency tables

library(flextable) # rendering tables

```

## Load data

```{r}

data<-read_sav("Dataset/HSE 2011.sav")

```

# Data exploration

## General dataset features
```{r}

colnames(data) #  variable names

dim(data) # number rows and columns

str(data) # general features of each variable

head(data) # first 6 rows 

```

The data includes 58 columns and 10617 rows (observations). The main variables of interest for subsequent analyses are:   
- HHSize (household size)
- Sex
- Age (age at last birthday)
- topqual3 (Highest Educational Qualification)
- totinc (Total Household Income)
- marstatc (Marital status including cohabitees)
- htval (Valid height (cm))
- wtval (Valid weight (Kg) inc. estimated>130kg)
- bmival (Valid BMI)
- dnnow (Whether drink nowadays)
- totalwu (total units of alcohol/week)
- gor1 (Government Office Region - numeric)



## Restrict dataset to variables of interest and rename them

```{r}
data%<>%
  select(age=Age,
         sex=Sex,
         household_size=HHSize,
         education=topqual3,
         household_income=totinc,
         marital_status=marstatc,
         height=htval,
         weight=wtval,
         bmi=bmival,
         drinks=dnnow,
         alcohol_units=totalwu,
         region=gor1)

```

# Descriptive statistics

### Age

```{r}
noquote("Summary statistics:")
summary(data$age)

noquote("Standard deviation:")
sd(data$age)

noquote(paste("Mode:", Mode(data$age)))


# calculate mode
noquote(paste0("Mode: ", round(with(density(data$age, na.rm = T), x[which.max(y)]),2))) ##   kernel density estimate

p1<-data%>%
  ggplot(aes(age))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")

p2<-data%>%
  ggplot(aes(age))+
  geom_density(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank())+
  labs(y="Density")
  
p3<-data%>%
  ggplot(aes(age))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  scale_x_continuous(breaks=seq(0,100,10))+
  labs(x="Age (at last birthday)")


  

p1/p2/p3+plot_layout(nrow = 3, heights = c(2, 1,1))->age_plot

age_plot

ggsave("Outputs/age_plot.png",
       width = 15,
       height=10,
       units="cm")

```


### Sex

```{r}

table(data$sex)

attr(data$sex, "labels")

data%>%
  ggplot(aes(as_factor(sex), fill=as_factor(sex)))+
  geom_text(stat="count",aes(label=paste0(after_stat(count), " (", round(after_stat(count)/length(data$sex)*100,1), "%)")),
            vjust=-0.2)+
  geom_bar()+
  labs(x="Sex",
       y="Count")+
  theme(legend.position = "none")->sex_plot

sex_plot

ggsave("Outputs/sex_plot.png")

```


### Marital status

```{r}
table(data$marital_status)

attr(data$marital_status, "labels")


data%>%
  ggplot(aes(as_factor(marital_status), fill=as_factor(marital_status)))+
  geom_bar()+
  geom_text(stat="count",aes(label=paste0(after_stat(count), "\n(", round(after_stat(count)/length(data$marital_status)*100,1), "%)")),
            vjust=-0.2)+
  
  labs(x="Marital status",
       y="Count")+
  theme(axis.text.x = element_text(angle=30, hjust=1,vjust=1),
        legend.position = "none")+
  scale_y_continuous(expand=expansion(c(0,0.5)))->marital_status_plot

marital_status_plot

ggsave("Outputs/marital_status_plot.png")


```


### Education
```{r}
table(as_factor(data$education))

attr(data$education, "labels")


data%>%
  ggplot(aes(as_factor(education), fill=as_factor(education)))+
  geom_bar()+
  geom_text(stat="count",aes(label=paste0(after_stat(count), "\n(", round(after_stat(count)/length(data$education)*100,1), "%)")),
            vjust=-0.2)+
  
  labs(x="Highest educational level achieved",
       y="Count")+
  theme(axis.text.x = element_text(angle=30, hjust=1,vjust=1, size=7),
        legend.position = "none",
        plot.margin = unit(c(0, 0, 0, 1),"cm"))+
  scale_y_continuous(expand=expansion(c(0,0.3)))->education_plot

education_plot

ggsave("Outputs/education_plot.png",
       width = 15,
       height=10,
       units = "cm")

```

### Height
```{r}

summary(data$height)


p1<-data%>%
  ggplot(aes(height))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")

p2<-data%>%
  ggplot(aes(height))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  labs(x="Height (in cm)")
  

p1/p2+plot_layout(nrow = 2, heights = c(2, 1))->height_plot

height_plot

ggsave("Outputs/height_plot.png")

```

### Weight
```{r}

summary(data$weight)


p1<-data%>%
  ggplot(aes(weight))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")

p2<-data%>%
  ggplot(aes(weight))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  labs(x="Weight (in kg)")
  

p1/p2+plot_layout(nrow = 2, heights = c(2, 1))->weight_plot

weight_plot

ggsave("Outputs/weight_plot.png")

```

### BMI
```{r}
noquote("Summary statistics:")
summary(data$bmi)

noquote(paste("Standard deviation:", sd(data$bmi, na.rm = T)))


# calculate mode
noquote(paste0("Mode: ", round(with(density(data$bmi, na.rm = T), x[which.max(y)]),2))) ##   kernel density estimate


p1<-data%>%
  ggplot(aes(bmi))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")+
  scale_x_continuous(breaks = seq(10,60,10))

p2<-data%>%
  ggplot(aes(bmi))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank())+
  scale_x_continuous(breaks = seq(10,60,10))

p3<-data%>%
  ggplot(aes(bmi))+
  geom_density(color="black")+
  scale_x_continuous(breaks = seq(10,60,10))+
  labs(y="Density",
       x="BMI")
  

p1/p2/p3+plot_layout(nrow = 3, heights = c(2, 1,1))->bmi_plot

bmi_plot

ggsave("Outputs/bmi_plot.png")



```

### Household size
```{r}

table(data$household_size)


noquote("Summary statistics:")
summary(data$household_size)

noquote(paste("Standard deviation:", sd(data$household_size, na.rm = T)))

data%>%
  ggplot(aes(household_size))+
  geom_bar(color="black")+
   geom_text(stat="count",aes(label=paste0(after_stat(count), "\n(", round(after_stat(count)/length(data$household_size)*100,1), "%)")),
            vjust=-0.2,
            size=3)+
  scale_x_continuous(limits=c(0,NA), breaks=seq(1,10,1))+
  labs(x="Household size",
       y="Count")+
    scale_y_continuous(expand=expansion(c(0,0.2)))->household_size_plot

household_size_plot

ggsave("Outputs/household_size_plot.png")


```

### Household income
```{r}

summary(data$household_income)


p1<-data%>%
  ggplot(aes(household_income))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")

p2<-data%>%
  ggplot(aes(household_income))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  labs(x="Household income")
  

p1/p2+plot_layout(nrow = 2, heights = c(2, 1))->household_income_plot

household_income_plot

ggsave("Outputs/household_income_plot.png")

```

### Region
```{r}


data%>%
  ggplot(aes(as_factor(region), fill=as_factor(region)))+
  geom_bar()+
  geom_text(stat="count",aes(label=paste0(after_stat(count), "\n(", round(after_stat(count)/length(data$region)*100,1), "%)")),
            vjust=-0.2)+
  
  labs(x="Region (within England)",
       y="Count")+
  theme(axis.text.x = element_text(angle=30, hjust=1,vjust=1),
        legend.position = "none")+
  scale_y_continuous(expand=expansion(c(0,0.3)))->region_plot

region_plot

ggsave("Outputs/region_plot.png")
```

### Current drinker
```{r}


data%>%
  ggplot(aes(as_factor(drinks), fill=as_factor(drinks)))+
  geom_text(stat="count",aes(label=paste0(after_stat(count), " (", round(after_stat(count)/length(data$drinks)*100,1), "%)")),
            vjust=-0.2)+
  geom_bar()+
  labs(x="Current drinker",
       y="Count")+
  theme(legend.position = "none")->drinks_plot

drinks_plot

ggsave("Outputs/drinks_plot.png")

```
## Units of alcohol per week

```{r}


summary(data$alcohol_units)


p1<-data%>%
  ggplot(aes(alcohol_units))+
  geom_histogram(color="black")+
  theme(axis.title.x = element_blank())+
  labs(y="Count")

p2<-data%>%
  ggplot(aes(alcohol_units))+
  geom_boxplot(color="black")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  labs(x="Alcohol units (per week)")
  

p1/p2+plot_layout(nrow = 2, heights = c(2, 1))->alcohol_units_plot

alcohol_units_plot

ggsave("Outputs/alcohol_units_plot.png")


```

# Inferential statistics

## Which gender drinks more alcohol

```{r}

# plot (barchart)
data%>%
  filter(!is.na(drinks))%>%
  count(as_factor(sex), as_factor(drinks))%>%
  group_by(`as_factor(sex)`)%>%
  mutate(prop=round(n/sum(n)*100,1))%>%
  rename(Sex=`as_factor(sex)`,
         `Current drinker` = `as_factor(drinks)`)%>%
  
  ggplot(aes(Sex, y=prop,fill=fct_rev(`Current drinker`)))+
  geom_col()+
  labs(x="Sex",
       y="Proportion",
       fill="Current drinker")+
  guides(fill = guide_legend(reverse = TRUE))+
  geom_text(data=.%>%filter(`Current drinker`=="Yes"), 
            aes(x=Sex, y=prop, label=paste0(n, " (", prop, "%)")),
            vjust=-0.1)+
  theme(legend.position = "bottom")

ggsave("Outputs/drinking_vs_sex.png",
       width=20,
       height=20,
       units="cm")

# contingency table

CrossTable(data$sex, 
           data$drinks,
           prop.c = F,
prop.r = T,
prop.t=F,
prop.chisq = F,
chisq = F,
format="SPSS")


# run test

chisq.test(data$drinks, data$sex)


```


## Which region drinks more alcohol

```{r}

# plot
data%>%
  filter(!is.na(drinks))%>%
  count(as_factor(region), as_factor(drinks))%>%
  group_by(`as_factor(region)`)%>%
  mutate(prop=round(n/sum(n)*100,1))%>%
  rename(Region=`as_factor(region)`,
         `Current drinker` = `as_factor(drinks)`)%>%
  
  ggplot(aes(Region, y=prop,fill=fct_rev(`Current drinker`)))+
  geom_col()+
  labs(x="Region",
       y="Proportion",
       fill="Current drinker")+
  guides(fill = guide_legend(reverse = TRUE))+
  geom_text(data=.%>%filter(`Current drinker`=="Yes"), 
            aes(x=Region, y=prop, label=paste0(n, "\n(", prop, "%)")),
            vjust=-0.1)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=30, hjust=1,vjust=1))

ggsave("Outputs/drinking_vs_region.png",
       width=20,
       height=20,
       units="cm")



# contingency table

CrossTable(data$region, 
           data$drinks,
           prop.c = F,
prop.r = T,
prop.t=F,
prop.chisq = F,
chisq = F,
format="SPSS")



# run test
chisq.test(data$drinks, data$region)



```

## Difference in height between men and women

```{r}

# Plot
p1<-data%>%
  select(height, sex)%>%
  ggplot(aes(height, group=as_factor(sex), fill=as_factor(sex)))+
  geom_histogram(alpha=0.5,
                 color="black",
                 bins=50)+
  labs(x="Height in cm (valid)",
       y="Count",
       fill="Sex")+
  theme(legend.position = "none",
        axis.title.x = element_blank())

p2<-data%>%
  select(height, sex)%>%
  ggplot(aes(height, y=as_factor(sex),group=as_factor(sex), fill=as_factor(sex)))+
  geom_boxplot(color="black")+
  labs(x="Height in cm (valid)",
       y="Sex",
       fill="Sex")+
  theme(legend.position = "bottom")

p1/p2

ggsave("Outputs/height_by_sex.png",
       width=20,
       height=20,
       units="cm")


# test for normality

ks.test(data$height, "pnorm")


# Table
data%>%
  select(height, sex)%>%
  filter(!is.na(height))%>%
  group_by(as_factor(sex))%>%
  summarise(Minimum=min(height),
            Mean=round(mean(height),1),
            SD = round(sd(height),1),
            Median=round(median(height),1),
            Q1 = round(quantile(height, .25),1),
            Q3 = round(quantile(height, .75),1),
            Maximum=round(max(height),1))%>%
  rename(Sex=`as_factor(sex)`)->height_by_sex_summary_table


height_by_sex_summary_table

write_csv(height_by_sex_summary_table, "Outputs/height_by_sex_summary_table.csv")

# Test

## non-parametric wilcoxon independent samples test
wilcox.test(height~sex, data)
       

```


## Difference in weight between men and women

```{r}



# Plot
p1<-data%>%
  select(weight, sex)%>%
  ggplot(aes(weight, group=as_factor(sex), fill=as_factor(sex)))+
  geom_histogram(alpha=0.5,
                 color="black",
                 bins=50)+
  labs(x="Weight in kg (valid)",
       y="Count",
       fill="Sex")+
  theme(legend.position = "none",
        axis.title.x = element_blank())

p2<-data%>%
  select(weight, sex)%>%
  ggplot(aes(weight, y=as_factor(sex),group=as_factor(sex), fill=as_factor(sex)))+
  geom_boxplot(color="black")+
  labs(x="Weight in cm (valid)",
       y="Sex",
       fill="Sex")+
  theme(legend.position = "bottom")

p1/p2

ggsave("Outputs/weight_by_sex.png",
       width=20,
       height=20,
       units="cm")


# test for normality

ks.test(data$weight, "pnorm")




# Table
data%>%
  select(weight, sex)%>%
  filter(!is.na(weight))%>%
  group_by(as_factor(sex))%>%
  summarise(Minimum=min(weight),
            Mean=round(mean(weight),1),
            SD = round(sd(weight),1),
            Median=round(median(weight),1),
            Q1 = round(quantile(weight, .25),1),
            Q3 = round(quantile(weight, .75),1),
            Maximum=round(max(weight),1))%>%
  rename(Sex=`as_factor(sex)`)->weight_by_sex_summary_table


weight_by_sex_summary_table

write_csv(weight_by_sex_summary_table, "Outputs/weight_by_sex_summary_table.csv")

# Test

## parametric independent samples  t-test 
t.test(weight~sex, data, var.equal=T)

## non-parametric wilcoxon independent samples test
wilcox.test(weight~sex, data)



```

## correlation between whether a person drinks nowadays, total household income, age at last birthday and gender?

```{r}

mycor<- function(x,...){
  r<- apply(x, 2, function(j){
    apply(x, 2, function(i){
      as.numeric(cor.test(i,j,...)$estimate)
    })
  })
  P<- apply(x, 2, function(j){
    apply(x, 2, function(i){
      as.numeric(cor.test(i,j,...)$p.value)
    })
  })
  out<-c()
  out$P<- P
  out$r<- r
  return(out) 
}

myCorDat<- mycor(data%>%
                   select(drinks, household_income, age, sex)%>%
                   mutate(drinks=case_when(drinks==1 ~ 1,
                                           drinks==2 ~ 0)), 
                 method="pearson", na.action=na.omit)

myCorDat



```